### 2021-02-25 abstract.tex
revision for jounal - abstract/introduction/background/study settings

\begin{abstract}
%  \emph{Context.}
%  The popularization of the Android platform and the growing number of Android apps that manage sensitive data turned the Android ecosystem into an attractive target for malicious software. For this reason, researchers and practitioners have investigated new approaches to address Android's security issues, including techniques that leverage dynamic analysis to mine Android Sandboxes. 
%  \emph{Problem.}
  The use of sandboxes is an effective technique for malware analysis. However, although the use of dynamic analysis for mining Android sandboxes has been investigated before, little is known about the potential benefits of combining static and dynamic analysis for mining Android sandboxes.
Accordingly, in this paper we present the results of two studies that investigate whether or not static analysis might complement and increase the performance of dynamic analysis tools for mining Android sandboxes. %\emph{Method.} 
In the first study we conduct a non-exact replication of the Bao et al. work, a previous study that compares the performance of test case generation tools for mining Android sandboxes. Differently from the original work, here we isolate the effect of the static analysis tool (DroidFax) they used to instrument the Android apps in their experiments. This decision was motivated by the fact that DroidFax could have influenced the efficacy of the dynamic analyses positively---through the execution of specific static analysis algorithms. In our second study, we carried out a new experiment to investigate the efficacy of tainted analysis algorithms for mining Android sandboxes. To this end, we executed the FlowDroid tool to mine the source-sink flows from the benign/malign pairs of Android apps used in the Bao et al. study. %\emph{Results.}
Our study brings several findings. For instance, the first study reveals that DroidFax alone (static analysis) can detect 43.75\% of the malwares in the dataset of the Bao et al. work, leading to an overestimation of the performance of the dynamic analysis tools. The results of the second study show that (static) tainted analysis is also practical for mining Android sandboxes, with a performance similar to that reached by dynamic analysis tools.
\end{abstract}

>>>>

\begin{abstract}
%  \emph{Context.}
The popularization of the Android platform and the growing number of Android apps that manage sensitive data turned the Android ecosystem into an attractive target for malicious software. For this reason, researchers and practitioners have investigated new approaches to address Android's security issues, including techniques that leverage dynamic analysis to \textit{mine Android sandboxes}. 
%  \emph{Problem.}
The use of sandboxes is an effective technique for malware analysis. However, although the use of dynamic analysis for mining Android sandboxes has been investigated before, little is known about the potential benefits of combining static and dynamic analysis for mining Android sandboxes.
Accordingly, in this paper we present the results of two studies that investigate whether or not static analysis might complement and increase the performance of dynamic analysis tools for mining Android sandboxes. %\emph{Method.} 
In the first study we conduct a non-exact replication of the Bao et al. work, a previous study that compares the performance of test case generation tools for mining Android sandboxes. Differently from the original work, here we isolate the effect of the static analysis tool (DroidFax) they used to instrument the Android apps in their experiments. This decision was motivated by the fact that DroidFax could have influenced the efficacy of the dynamic analyses positively---through the execution of specific static analysis algorithms. In our second study, we carried out a new experiment to investigate the efficacy of tainted analysis algorithms for mining Android sandboxes. To this end, we executed the FlowDroid tool to mine the source-sink flows from the benign/malign pairs of Android apps used in the Bao et al. study. %\emph{Results.}
Our study brings several findings. For instance, the first study reveals that DroidFax alone (static analysis) can detect 43.75\% of the malwares in the dataset of the Bao et al. work, leading to an overestimation of the performance of the dynamic analysis tools. The results of the second study show that (static) tainted analysis is also practical for mining Android sandboxes, with a performance similar to that reached by dynamic analysis tools.
\end{abstract}

### 2021-02-25 background.tex
revision for jounal - abstract/introduction/background/study settings

L Bao et al~.\cite{DBLP:conf/wcre/BaoLL18} had provided an important comparative study between five test generation tools that find malware by mining sandboxes techniques: DroidMate, Monkey, GUIRipper \cite{DBLP:conf/kbse/AmalfitanoFTCM12}, Puma \cite{DBLP:conf/mobisys/Hao0NHG14}, and Droidbot. This study indicates that these tools was efficient in identifying at most 70\% of the malware in a specific dataset and also reports that after combining all test generator tools, it was possible to detect 75.49\% of malicious behavior explored ($77$ among $102$). However, this study did not focus on the possible interference of static analysis in the final result, since this study used a instrumentation tool \cite{DBLP:conf/icsm/CaiR17a}, that also performs a static analysis of the apps

>>>>

L Bao et al~.\cite{DBLP:conf/wcre/BaoLL18} had provided an important comparative study between five test generation tools that find malware by mining sandboxes techniques: DroidMate, Monkey, GUIRipper \cite{DBLP:conf/kbse/AmalfitanoFTCM12}, Puma \cite{DBLP:conf/mobisys/Hao0NHG14}, and Droidbot. This study indicates that these tools was efficient in identifying at most $70$\% of the malware in a specific dataset and also reports that after combining all test generator tools, it was possible to detect $75.49$\% of malicious behavior explored ($77$ among $102$). However, this study did not focus on the possible interference of static analysis in the final result, since this study used a instrumentation tool \cite{DBLP:conf/icsm/CaiR17a}, that also performs a static analysis of the apps

### 2021-02-25 introduction.tex
revision for jounal - abstract/introduction/background/study settings

Almost two-thirds of the world population use mobile technologies~\cite{Comscore}, and the Android Operating System has dominated the market of smartphones, tablets, and others electronic devices \cite{statcounter}. Due to this growing popularity, the number of incidents related to Android malicious software (malware) has significantly increased~\cite{DBLP:journals/comsur/FarukiBLGGCR15,DBLP:journals/csur/SufatrioTCT15}. Security issues in Android software applications~\footnote{In this paper, we will use the terms Android Applications, Android Apps and Apps interchangeably to represent Android software applications} have become a relevant research topic, and many techniques have been developed to identify vulnerabilities in Android apps \cite{DBLP:conf/pldi/ArztRFBBKTOM14}, including the use of static analysis algorithms to identify privacy leaks, and to reveal the misuse of cryptographic primitives~\cite{krueger:ecoop-2018,rahaman:ccs-2019}, for instance. Another approach explored by researchers to mitigate the same problem is the use of dynamic analysis to mine Android sandboxes. In this proposal, given an app A, a sandbox S identifies sensitive resources access by A, during dynamic analysis, and then block future calls to sensitive resources unidentified during testing, thereby protecting Android users from malicious behavior.~\cite{DBLP:conf/icse/JamrozikSZ16}.

The main idea of mining sandboxes is to explore the set of calls to sensitive API methods in a Android app, using different approaches for automatically testing Android apps. These analysis explore apps behavior and record all calls to sensitive APIs. A sandbox builds upon the calls to sensitive Android APIs, during the execution of the test cases (exploratory phase). The sandbox could then block future calls to other sensitive resources which diverge from those found in the exploratory phase. Using this approach, the sandbox accuracy depends on the exploratory capabilities of each testing tool used to mine the sandbox rules.

Jamrozik et al.~\cite{DBLP:conf/icse/JamrozikSZ16} argue in favor of dynamic analysis for mining sandboxes, instead of using static analysis. The authors claim that dynamic analysis exceeds static analysis on mining Android sandboxes because the analysis must often assume that additional behaviors are possible than actually would be \kn{Ok. In the first paragraph we discuss static analysis in general to detect vulnerabilities and now without mentioning it prior we seem to mention static analysis can also mine sandboxes. The overall flow of definitions needs some thought}. Otherwise, static analysis might determine that some behavior is impossible and therefore no longer needs to be checked at runtime. Thereby, code that is interpreted, decrypted, or downloaded at run time is out of scope for static analysis---even though these more \emph{dynamic} strategies are often used to introduce malicious behavior. Nonetheless, the authors do not give any empirical validation about the limitations of static analysis to mine sandboxes. As a consequence, it is not clear whether and how both approaches (dynamic and static analysis) could complement each other in the process of mining Android sandboxes.

@@

Our goal is to understand how static analysis
algorithms might complement dynamic analysis approaches in the process of mining Android sandboxes. To this end, we performed two studies. In the
first, we conduct a non-exact replication of the work of \blls. Our study differs from the original work in three aspects: (a)
by isolating the effect of the static analysis component of DroidFax, (b) by considering more recent dynamic analysis tools for testing Android apps,
and (c) we considering more execution time for each apps analysed at our replication. In the second study, we explore to what extent we can complement the dynamic analysis approach using tainted analysis algorithms, through Flowdroid \cite{DBLP:conf/pldi/ArztRFBBKTOM14} tool.

Altogether, this paper brings the following contributions:

\todo[inline]{RB.: Later we can enumerate the paper contributions here.}
The rest of the paper is organized as follows. Section 2 we present a background and some related works. Section 3 introduces the study setting of our research and Section 4 complements presenting some results and discussion, highlighting motivation malware examples and the use of a promising static analysis tool, on these examples. Finally, Sections 5 and 6, summarizes the paper, presenting some threats, presenting the final conclusions, and discussing our ideas for future works.

>>>>

Almost two-thirds of the world population use mobile technologies~\cite{Comscore}, and the Android Operating System has dominated the market of smartphones, tablets, and others electronic devices \cite{statcounter}. Due to this growing popularity, the number of incidents related to Android malicious software (malware) has significantly increased~\cite{DBLP:journals/comsur/FarukiBLGGCR15,DBLP:journals/csur/SufatrioTCT15}. Security issues in Android software applications~\footnote{In this paper, we will use the terms Android Applications, Android Apps and Apps interchangeably to represent Android software applications} have become a relevant research topic, and many techniques have been developed to identify vulnerabilities in Android apps~\cite{DBLP:conf/pldi/ArztRFBBKTOM14}, including the use of static analysis algorithms to identify privacy leaks, and to reveal the misuse of cryptographic primitives~\cite{krueger:ecoop-2018,rahaman:ccs-2019}, for instance. Another approach explored by researchers to mitigate the same problem is the use of dynamic analysis to mine Android sandboxes. In this proposal, given an app A, a sandbox S identifies sensitive resources access by A, during dynamic and static analysis, and then block future calls to sensitive resources unidentified during this exploratory phase, thereby protecting Android users from malicious behavior.~\cite{DBLP:conf/icse/JamrozikSZ16}.

The main idea of mining sandboxes is to explore the set of calls to sensitive API methods in a Android app, using different approaches for automatically testing Android apps. These analysis explore apps behavior and record all calls to sensitive APIs. A sandbox builds upon the calls to sensitive Android APIs, during the execution of the test cases (exploratory phase). The sandbox could then block future calls to other sensitive resources which diverge from those found in the exploratory phase. Using this approach, based on just test case generation tools, the sandbox accuracy depends on the exploratory capabilities of each testing tool used to mine the sandbox security rules.

Jamrozik et al.~\cite{DBLP:conf/icse/JamrozikSZ16} argue in favor of dynamic analysis for mining sandboxes, instead of using static analysis. The authors claim that dynamic analysis exceeds static analysis on mining Android sandboxes because the analysis must often assume that additional behaviors are possible than actually would be.
%\kn{Ok. In the first paragraph we discuss static analysis in general to detect vulnerabilities and now without mentioning it prior we seem to mention static analysis can also mine sandboxes. The overall flow of definitions needs some thought}
Otherwise, static analysis might determine that some behavior is impossible and therefore no longer needs to be checked at runtime. Thereby, code that is interpreted, decrypted, or downloaded at run time is out of scope for static analysis---even though these more \emph{dynamic} strategies are often used to introduce malicious behavior. Nonetheless, the authors do not give any empirical validation about the limitations of static analysis to mine sandboxes. As a consequence, it is not clear whether and how both approaches (dynamic and static analysis) could complement each other in the process of mining Android sandboxes.

@@

Our goal is to understand how static analysis
algorithms might complement dynamic analysis approaches in the process of mining Android sandboxes. To this end, we performed two studies. In the
first, we conduct a replication of the work of \blls. Our study differs from the original work in three aspects: (a)
by isolating the effect of the static analysis component of DroidFax, (b) by considering more recent dynamic analysis tools for testing Android apps,
and (c) we considering more execution time, and more repetitions, for each apps analysed at our replication. In the second study, we explore to what extent we can complement the dynamic analysis approach using tainted analysis algorithms, through Flowdroid \cite{DBLP:conf/pldi/ArztRFBBKTOM14} tool.

\textit{Extended Version.} This journal paper extends a previous conference publication~\cite{DBLP:conf/scam/CostaMCMVBC20} by expanding the empirical evaluation of sandboxes solutions using DroidXP benchmark tool with:
 \begin{enumerate}[(1)]
  \item presenting a replication of the accuracy study of test generation tools for mining android sandboxes, using an extension and update of our benchmark solution, that aided us disables the static analysis performed by DroidFax.

  \item discovering which are the contribution installments of types of analysis for malware detection task.

  \item presenting what malware and its characteristics, that was not detected by any tool in our analysis dynamic and static experiment.

  \item presenting data that indicate that static analysis can improve the mining sandboxes process for malware detection.
  \end{enumerate}

%Altogether, this paper brings the following contributions:
%\todo[inline]{RB.: Later we can enumerate the paper contributions here.}
The rest of the paper is organized as follows. Section 2 we present a background and some related works. Section 3 introduces the study setting of our research and Section 4 complements presenting some results and discussion, highlighting motivation malware examples and the use of a promising static analysis tool, on these examples. Finally, Sections 5 and 6, summarizes the paper, presenting some threats, presenting the final conclusions, and discussing our ideas for future works.

### 2021-02-25 setup.tex
revision for jounal - abstract/introduction/background/study settings

\blls explored the effectiveness of test generation tools to construct sandboxes, based on sensitive APIs called by apps under analysis. Since their study also involved a static analysis component, performed by Droidfax as we described at introduction, and whose impact was not measured, we perform a non-exact replication to understand the static analysis impact may have had in the results. Thus, this section present the settings of our study, whose goal is to perform this non-exact replication, and thus build a general understanding of the implications of static analysis algorithms in their results. We also investigate how static analysis can improve the performance of mining sandboxes, in the task of identifying malicious behavior.

@@

We conducted two studies to answer the research questions above. First we performed a non-exact replication of the \blls. Our study differs from the original because here we isolated the effect of the DroidFax static analysis algorithms in the performance to identify malicious apps using the test case generation tools for
mining sandboxes. In addition, we discarded $6$ pairs of
Android apps we were not able to instrument---out of the $102$ pairs used in the original work. We also introduced a recent test generator tool, that has not been considered at previous work, Humanoid ~\cite{DBLP:conf/kbse/LiY0C19}, and different from the original study, we expand the execution time of each test generation tool, executing each app at each tool for three minutes, while the original study executed for just one minute.

The others $3$ test generation tools selected were: Droidbot~\cite{DBLP:conf/icse/LiYGC17},
DroidMate~\cite{DBLP:conf/icse/JamrozikZ16} and Monkey~\cite{Monkey}, all explored at \blls. We selected Droidbot and DroiMate because they achieved
the best performance on detecting malicious behavior at $102$ pairs of Android apps (B/M) in the \blls. We also considered the open source tool from Google, Monkey, because it is the most widely used tool in industry \cite{DBLP:conf/sigsoft/ZengLZXDLYX16}. It is part of the Android SDK, and does not require any additional installation effort. Since Monkey was developed for stress testing, and can not perform text input, only generating
UI events, we also considered Humanoid, that emulate realistic users, creating human-like test inputs using deep learning techniques.
For this non-exact replication, we also used the DroidXP benchmark~\cite{DBLP:conf/scam/CostaMCMVBC20},
which helped us to reproduce the original work.

In the second study we leverage FlowDroid~\cite{DBLP:conf/pldi/ArztRFBBKTOM14} to execute
tainted analysis algorithms in Android apps, in order to identify pairs of \emph{sources and sinks}. In this case,
our goal is to investigate the performance on detecting malicious
behavior using static taint analysis of Android apps. We use two metrics in this second study: the number
of source-sink flows that FlowDroid identifies when considering the pairs of Android apps (B/M) and the
execution time of the analysis for each app.
For both studies, we employed the same dataset $96$ pairs of Android apps (B/M),
shared by the AndroZoo \cite{DBLP:conf/msr/AllixBKT16} project. We detail the procedures of each study in what follows.

\subsection{First Study: A non-exact replication}

@@

We investigate the four test case generation tools described earlier and added a fake test
case generation tool (named Joker) that simulates a test tool that does not execute
the Android apps during a benchmark execution. Using this tool, the results
of the dynamic analysis are not considered and we can compute the results with
only the static analysis component of DroidFax (RQ1). Our study executed each pairs of
Android app (B/M) in each one of the five test generation tools, including Joker,
for $3$ minutes. We then investigate the performance of each tool to detect
malicious behaviors in the sandbox modeled by the test generation tool
under analysis. Figure \ref{fig:setup} shows this experimental setup.

@@

After that, we replicated the study using a configuration of DroidXP that
disables the DroidFax static analysis algorithm. With this new configuration,
we executed the benchmark one more time, using the same dataset of Android
apps, the same execution time, and the same test case generation tools.
The results of this replication allow us to compute the effective performance
of the dynamic analysis tools (RQ2)---that is, ignoring the influence of the
DroidFax static analysis algorithms.

>>>>

\blls explored the effectiveness of test generation tools to construct sandboxes, based on sensitive APIs called by apps under analysis. Since their study also involved a static analysis component, performed by Droidfax as we described at introduction, and whose impact was not measured, we replicated its work to understand the static analysis impact may have had in their results. Thus, this section present the settings of our study, whose goal is to perform this replication, and thus build a general understanding of the implications of static analysis algorithms in the results. We also investigate how static analysis can improve the performance of mining sandboxes, in the task of identifying malicious behavior.

@@

We conducted two studies to answer the research questions above. First we replicated the \blls. However, our study differs from the original because here we isolated the effect of the DroidFax static analysis algorithms, in the task to identify malicious apps. In addition, we discarded $6$ pairs of
Android apps we were not able to instrument---out of the $102$ pairs used in the original work. We also introduced a recent test generator tool, that has not been considered at previous work, Humanoid ~\cite{DBLP:conf/kbse/LiY0C19}, and different from the original study, we expand the execution time of each test generation tool, executing each app at each tool for three minutes, and repeated all the execution more 2 times, therefore totaling three executions. The original study executed each app at each tool, for just one minute, and just one time.

The others $3$ test generation tools selected were: Droidbot~\cite{DBLP:conf/icse/LiYGC17},
DroidMate~\cite{DBLP:conf/icse/JamrozikZ16} and Monkey~\cite{Monkey}, all explored at \blls. We selected Droidbot and DroiMate because they achieved
the best performance on detecting malicious behavior at $102$ pairs of Android apps (B/M) in the \blls. We also considered the open source tool from Google, Monkey, because it is the most widely used tool in industry \cite{DBLP:conf/sigsoft/ZengLZXDLYX16}. It is part of the Android SDK, and does not require any additional installation effort. Since Monkey was developed for stress testing, and can not perform text input, only generating
UI events, we also considered Humanoid, that emulate realistic users, creating human-like test inputs using deep learning techniques.
For the replication, we also used our benchmark tool, DroidXP ~\cite{DBLP:conf/scam/CostaMCMVBC20},
which helped us to reproduce the original work.

In the second study we leverage FlowDroid~\cite{DBLP:conf/pldi/ArztRFBBKTOM14} to execute
tainted analysis algorithms in Android apps, in order to identify pairs of \emph{sources and sinks}. In this case,
our goal is to investigate the performance on detecting malicious
behavior using static taint analysis of Android apps. We use two metrics in this second study: the number
of source-sink flows that FlowDroid identifies when considering the pairs of Android apps (B/M), and the
execution time of the analysis for each app.
For both studies, we employed the same dataset $96$ pairs of Android apps (B/M),
shared by the AndroZoo \cite{DBLP:conf/msr/AllixBKT16} project. We detail the procedures of each study in what follows.

\subsection{First Study: A replication \blls}

@@

We investigate the four test case generation tools described earlier and added a fake test
case generation tool (named Joker) that simulates a test tool that does not execute
the Android apps during a benchmark execution. Using this tool, the results
of the dynamic analysis are not considered and we can compute the results with
only the static analysis component of DroidFax (RQ1). Our study executed each pairs of
Android app (B/M) in each one of the five test generation tools, including Joker,
for three minutes, and for three times. We then investigate the performance of each tool to detect malicious behaviors in the sandbox modeled by the test generation tool
under analysis. Figure \ref{fig:setup} shows this experimental setup.

@@

After that, we replicated the study using a configuration of DroidXP that disables the DroidFax static analysis algorithm. With this new configuration, we executed the benchmark again, using the same dataset of Android pairs apps, the same execution time, and the same test case generation tools.
The results of this replication allow us to compute the effective performance
of the dynamic analysis tools (RQ2)---that is, ignoring the influence of the
DroidFax static analysis algorithms.
